\documentclass[11pt]{article}

% Wiley-style formatting for Journal of Software: Evolution and Process
\usepackage[margin=1in]{geometry}
\usepackage{times} % Wiley uses Times font

% Essential packages
\usepackage[numbers,sort&compress]{natbib}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\newtheorem{definition}{Definition}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{url}
\usepackage[english]{babel}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}

\usepackage{float}
\floatstyle{plain}
\newfloat{listing}{htbp}{lop}
\floatname{listing}{Listing}

\usepackage{multirow}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, positioning, fit, backgrounds}
\usepackage{pgfplots}
\pgfplotsset{compat=1.17}
\usepackage{subcaption}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}

% Line numbers for review (uncomment for submission)
% \usepackage{lineno}

% Wiley-style section formatting
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalsize\itshape}{\thesubsubsection}{1em}{}

\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codeblue}{rgb}{0.0,0.0,0.6}
\definecolor{codegreen}{rgb}{0,0.5,0}
\definecolor{lightblue}{RGB}{173,216,230}
\definecolor{lightgreen}{RGB}{144,238,144}
\definecolor{lightyellow}{RGB}{255,255,224}
\definecolor{lightpurple}{RGB}{221,160,221}

\lstdefinestyle{code}{
  language=Java,
  basicstyle=\fontsize{9}{11}\selectfont\ttfamily,
  numbers=none,
  numberstyle=\tiny\color{codegray},
  stepnumber=1,
  numbersep=5pt,
  backgroundcolor=\color{white},
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  frame=single,
  rulecolor=\color{black},
  tabsize=2,
  captionpos=b,
  breaklines=true,
  breakatwhitespace=true,
  title=\lstname,
  keywordstyle=\color{codeblue},
  commentstyle=\color{codegreen}\itshape,
  stringstyle=\color{red},
  escapeinside={(*@}{@*)},
  morekeywords={*,...}
}

\lstset{style=code}

\usepackage{textcomp}

% Running headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\itshape Edmonds and Utting}
\fancyhead[R]{\small\itshape J. Softw. Evol. Process}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\begin{document}

% Enable line numbers for review (uncomment for submission)
% \linenumbers

% Article type
\noindent\textbf{RESEARCH ARTICLE}
\vspace{1em}

% Title
\begin{center}
{\LARGE\bfseries Automated Formal Specification Inference for Enhanced Program Comprehension and Test Generation: A Static Analysis Approach}
\vspace{1.5em}

% Authors
{\large Brendan Edmonds\textsuperscript{1}* \quad Mark Utting\textsuperscript{1}}
\vspace{0.5em}

% Affiliations
{\small\textsuperscript{1}UQ Cyber, School of Electrical Engineering and Computer Science, University of Queensland, St Lucia QLD 4067, Australia}
\vspace{0.5em}

% Correspondence
{\small *Correspondence: Brendan Edmonds, b.edmonds@uq.edu.au}
\end{center}

\vspace{1em}

% Abstract
\noindent\textbf{Abstract}
\vspace{0.5em}

\noindent Understanding software behavior is critical for maintenance and evolution, especially when documentation is incomplete. This paper presents JML-Inferrer, a static analysis tool that automatically infers formal specifications from Java methods using weakest precondition (WP) and strongest postcondition (SP) reasoning. The tool derives precise method contracts that facilitate program comprehension, automated test generation, and interface validation.

We describe the tool's architecture, including WP/SP interaction, loop invariant inference heuristics, and handling of complex constructs. Applied to 10,709 methods from the OpenJDK core library, JML-Inferrer achieves 94.2\% precision and 89.3\% recall for preconditions, and 87.6\% precision and 78.1\% recall for postconditions, validated through manual inspection of 500 samples.

Using specifications to guide Large Language Model (LLM)-based test generation (Gemini 2.0, temperature 0.3, 5 runs), we observe a 68.4\% improvement in test count (95\% CI: [62.1\%, 74.7\%]) and a 34.1 percentage point improvement in mutation score over signature-only baselines. Notably, specifications outperform source-code-based generation by 3.5 percentage points in mutation score, demonstrating that formal contracts provide more actionable guidance than raw implementation details.

A complete replication package is available, including source code, datasets, and analysis scripts. This work bridges formal methods and practical software engineering through lightweight, automated specification inference.

\vspace{1em}

% Keywords
\noindent\textbf{Keywords:} specification inference; weakest preconditions; strongest postconditions; program comprehension; software evolution; LLM-based testing; static analysis; mutation testing

\vspace{1em}
\hrule
\vspace{1em}

\input{introduction_journal}
\input{background_journal}
\input{approach_journal}
\input{methodology_journal}
\input{results_journal}
\input{discussion_journal}
\input{threats_journal}
\input{related_work_journal}
\input{conclusion_journal}

\section*{Data Availability Statement}

A complete replication package is available at:\\
\url{https://github.com/[anonymized]/jml-inferrer}

The package includes:
\begin{itemize}
    \item \textbf{Tool source code}: Complete Java implementation with build instructions
    \item \textbf{Method dataset}: Full list of 10,709 OpenJDK methods with selection criteria and categorization
    \item \textbf{Inferred specifications}: All generated JML specifications in machine-readable format
    \item \textbf{Test suites}: Generated tests for all three experimental phases (P1, P2, P3)
    \item \textbf{Raw results}: Mutation testing logs and statistical analysis scripts
    \item \textbf{Manual validation data}: Annotated sample of 500 specifications with correctness labels
\end{itemize}

\section*{Acknowledgments}

This research was supported by the University of Queensland Cyber initiative. We thank the anonymous reviewers for their constructive feedback.

\section*{Conflict of Interest}

The authors declare no conflict of interest.

\section*{Author Contributions}

\textbf{Brendan Edmonds}: Conceptualization, Methodology, Software, Validation, Investigation, Data Curation, Writing - Original Draft, Visualization. \textbf{Mark Utting}: Conceptualization, Methodology, Writing - Review \& Editing, Supervision.

\section*{ORCID}

Brendan Edmonds \url{https://orcid.org/0000-0000-0000-0000}\\
Mark Utting \url{https://orcid.org/0000-0000-0000-0000}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
