\documentclass[10pt,conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.


\usepackage[numbers]{natbib}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{url} 
\usepackage[english]{babel}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}



\usepackage{float}
\floatstyle{plain}
\newfloat{listing}{htbp}{lop}
\floatname{listing}{Listing}

\usepackage{multirow}
\usepackage{listings}
\usepackage{xcolor} % for custom colors

\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codeblue}{rgb}{0.0,0.0,0.6}
\definecolor{codegreen}{rgb}{0,0.5,0}

\lstdefinestyle{IEEE}{
  language=Java, % Change to your language (e.g., Java, C++)
  basicstyle=\fontsize{10}{10}\selectfont\ttfamily,
  numbers=none,
  numberstyle=\tiny\color{codegray},
  stepnumber=1,
  numbersep=5pt,
  backgroundcolor=\color{white},
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  frame=single,
  rulecolor=\color{black},
  tabsize=2,
  captionpos=b,
  breaklines=true,
  breakatwhitespace=true,
  title=\lstname,
  keywordstyle=\color{codeblue},
  commentstyle=\color{codegreen}\itshape,
  stringstyle=\color{red},
  escapeinside={(*@}{@*)}, % for LaTeX in listings
  morekeywords={*,...} % Add custom keywords if needed
}

\lstset{style=IEEE}

\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Enhancing Program Comprehension and Test Generation via Formal Specification Inference}

%\author{\IEEEauthorblockN{1\textsuperscript{st} Anonymous Authors}}
\author{\IEEEauthorblockN{1\textsuperscript{st} Brendan Edmonds}
\IEEEauthorblockA{\textit{UQ Cyber, University of Queensland} \\
\textit{University of Queensland)}\\
St Lucia QLD 4067, Australia \\
b.edmonds@uq.edu.au}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Mark Utting}
\IEEEauthorblockA{\textit{UQ Cyber, University of Queensland} \\
\textit{University of Queensland)}\\
St Lucia QLD 4067, Australia \\
m.utting@uq.edu.au}
}


\maketitle

\begin{abstract}
Understanding and validating the behaviour of shared components in systems is critical, especially when documentation is incomplete or source code is unavailable. This paper presents a static analysis technique that automatically infers formal specifications from Java methods using weakest precondition (WP) and strongest postcondition (SP) reasoning. The approach derives precise method contracts—normal-behaviour preconditions and postconditions—and embeds them in compiled class files, with the long-term goal of facilitating improved testing of library clients.

The inferred specifications enhance program comprehension, guide automated test generation, and support long-term software maintenance, particularly in environments where APIs evolve independently. Applied to the OpenJDK core library, our tool enables AI-based unit test generation to produce up to twice as many tests as baseline methods and achieves mutation scores up to 95.7\% across method categories including accessors, factory methods, and control-flow structures. Embedding these specifications within shared libraries strengthens test completeness in safety- or compliance-critical workflows and has the potential to enable interface validation across microservices.

Overall, this work bridges formal methods and practical software engineering by showing how lightweight, inference-based specifications can improve the reliability, maintainability, and evolution of distributed systems.
\end{abstract}


\begin{IEEEkeywords}
Specification Inference, Weakest Preconditions, Program Comprehension, LLM Testing
\end{IEEEkeywords}

\input{introduction}
\input{method}
\input{results}
\input{discussion}
\input{conclusion}

\section*{Data Availability Statement}

The data supporting the findings of this study are not publicly available due to reproducibility constraints. The experimental results depend on outputs generated by large language models (LLMs), whose behavior may vary across model versions or updates. As such, identical reproduction of the AI-generated data cannot be guaranteed. However, all analysis procedures and configurations used in this study are available from the authors upon reasonable request.

\input{references}

\end{document}
